---
title: "ISE 539: Special Credit Problem"
author: "Yan Xuan 5762411409"
date: "April 6, 2021"
output:
  html_document:
    df_print: paged
  pdf_document: default
---
----------------------------------------------------------------------------------------------------------------
              
         
Initialize p<sub>i</sub> for each element i, where i=1,2,...10.
```{r}
n=100
set.seed(1)
u1=runif(10)
p=u1/sum(u1)
elements=seq(1:10)
p_order=sort(p,decreasing=TRUE)
elements_order=order(p,decreasing=TRUE)
list_initial=sample(1:10,10,replace=FALSE)
```

Functions of list problem when reordering is not applied:
```{r}
list_problem_direct<-function(p=p, elements=elements,list_initial=list_initial, 
                              rule="interchange",var_reduce="control"){
  u2=runif(100)
  u2_anti=1-u2
  X=0
  X_anti=0
  Y=sum(u2)
  if (rule=="interchange"){
    list_interchange=list_initial
    for (i in 1:100){
      p_reordered=p[list_interchange]
      p_reordered_cum=cumsum(p_reordered)
      element_request=list_interchange[which(u2[i]<=p_reordered_cum)[1]]
      position_request=which(list_interchange==element_request)
      #print(u2[i])
      #print(p_order_cum)
      #print(position_request)
      #print(element_request)
      if (position_request!=1){
        list_interchange[position_request]=list_interchange[position_request-1]
        list_interchange[position_request-1]=element_request
      }
      #print(list_interchange)
      X=X+position_request
    }
    if (var_reduce=="control"){
      result=c(X,Y)
      return(result)
    }
    else if(var_reduce=="antithetic"){
      list_interchange=list_initial
      for (i in 1:100){
        p_reordered=p[list_interchange]
        p_reordered_cum=cumsum(p_reordered)
        element_request=list_interchange[which(u2_anti[i]<=p_reordered_cum)[1]]
        position_request=which(list_interchange==element_request)
        if (position_request!=1){
          list_interchange[position_request]=list_interchange[position_request-1]
          list_interchange[position_request-1]=element_request
        }
        X_anti=X_anti+position_request
      }
      result=c(X,X_anti)
      return(result)
    }
  }
  else if (rule=="front"){
    list_front=list_initial
    for (i in 1:100){
      p_reordered=p[list_front]
      p_reordered_cum=cumsum(p_reordered)
      element_request=list_front[which(u2[i]<=p_reordered_cum)[1]]
      position_request=which(list_front==element_request)
      if (position_request!=1){
        list_front=c(list_front[position_request],list_front[-position_request])
      }
      X=X+position_request
    }
    if (var_reduce=="control"){
      result=c(X,Y)
      return(result)
    }
    else if (var_reduce=="antithetic"){
      list_front=list_initial
      for (i in 1:100){
        p_reordered=p[list_front]
        p_reordered_cum=cumsum(p_reordered)
        element_request=list_front[which(u2_anti[i]<=p_reordered_cum)[1]]
        position_request=which(list_front==element_request)
        if (position_request!=1){
          list_front=c(list_front[position_request],list_front[-position_request])
        }
        X_anti=X_anti+position_request
      }
      result=c(X,X_anti)
      return(result)
    }
  }
}
```

Function of list problem when reordering is applied:
```{r}
list_problem_reorder<-function(p=p_order, elements=elements_order,list_initial=list_initial,
                               rule="interchange",var_reduce="control"){
  p_cum=cumsum(p)
  u2=runif(100)
  u2_anti=1-u2
  X=0
  X_anti=0
  Y=sum(u2)
  if (rule=="interchange"){
    list_interchange=list_initial
    for (i in 1:100){
      element_request=elements[which(u2[i]<=p_cum)[1]]
      position_request=which(list_interchange==element_request)
      if (position_request!=1){
        list_interchange[position_request]=list_interchange[position_request-1]
        list_interchange[position_request-1]=element_request
      }
      #print(element_request)
      #print(list_interchange)
      
      X=X+position_request
    }
    if (var_reduce=="control"){
      result=c(X,Y)
      return(result)
    }
    else if(var_reduce=="antithetic"){
      list_interchange=list_initial
      for (i in 1:100){
        element_request=elements[which(u2_anti[i]<=p_cum)[1]]
        position_request=which(list_interchange==element_request)
        if (position_request!=1){
          list_interchange[position_request]=list_interchange[position_request-1]
          list_interchange[position_request-1]=element_request
        }
        X_anti=X_anti+position_request
      }
      result=c(X,X_anti)
      return(result)
    }
  }
  else if (rule=="front"){
    list_front=list_initial
    for (i in 1:100){
      element_request=elements[which(u2[i]<=p_cum)[1]]
      position_request=which(list_front==element_request)
      if (position_request!=1){
        list_front=c(list_front[position_request],list_front[-position_request])
      }
      #print(element_request)
      #print(list_front)
      
      X=X+position_request
    }
    if (var_reduce=="control"){
      result=c(X,Y)
      return(result)
    }
    else if (var_reduce=="antithetic"){
      list_front=list_initial
      for (i in 1:100){
        element_request=elements[which(u2_anti[i]<=p_cum)[1]]
        position_request=which(list_front==element_request)
        if (position_request!=1){
          list_front=c(list_front[position_request],list_front[-position_request])
        }
        X_anti=X_anti+position_request
      }
      result=c(X,X_anti)
      return(result)
    }
  }
}
```

### Problem 1
Consider the list problem (Ex 9i) . Let n=10. Let V_1, ..., V_{10} be indep uniform (0, 1)  and set p_i = V_i / sum_j V_j.

Let X be the sum of the positions of the first 100 elements requested (N=100).   Use the method in the text to determine  in each period  which element is requested. Let U_j, j=1,...,100  be the random number used to determine which element is requested in period j . Suppose the **interchange rule** is used

(a) Estimate E[X]  based on 10,000 runs, using  Y = sum U_j  as a control variable (its mean is 50) .   Also estimate the variance of the estimator  X + c*(Y - 50) and compare it to estimate of Var(X) 

(b) Estimate E[X] based on 5.000 runs using antithetic variables. In each run the estimator is    E= (X(U_1,...,U_{100}) + X( 1- U_1, ...,1-U_{100}))/2.  estimate the variance of E

(c) Which of control variable or  antithetic gives greater variance reduction.

#### 1. (a)
```{r}
X={}
Y={}
for (j in 1:n){
  result=list_problem_direct(p=p, elements=elements,list_initial=list_initial,
                             rule="interchange",var_reduce="control")
  X=c(X,result[1])
  Y=c(Y,result[2])
}
estimator_raw=X #raw estimator
cat(paste('Mean of raw estimator: ', mean(estimator_raw), 
          '\nVariance of raw estimator: ',var(estimator_raw)))
```
Using **control variables** method:
```{r}
c=-cov(X,Y)/var(Y)
estimator_control=X+c*(Y-50) #estimator using interchange rule, control variable method
cat(paste('Mean of estimator using control variables: ', mean(estimator_control), 
          '\nVariance of estimator using control variables: ',var(estimator_control)))
```
#### 1. (b)
Using **antithetic variables** method:
```{r}
X={}
X_anti={}
for (j in 1:(n/2)){
  result=list_problem_direct(p=p, elements=elements,list_initial=list_initial,
                             rule="interchange",var_reduce="antithetic")
  X=c(X,result[1])
  X_anti=c(X_anti,result[2])
}
estimator_anti=(X+X_anti)/2 #estimator using interchange rule, antithetic variable method
cat(paste('Mean of estimator using antithetic variables: ', mean(estimator_anti), 
          '\nVariance of estimator using antithetic variables: ',var(estimator_anti)))
```
#### 1. (c)
```{r}
cat(paste('Variance reduction using control variables: ', paste(round(100*(var(estimator_raw)-var(estimator_control))/var(estimator_raw),2),"%", sep = "" ), 
          '\nVariance reduction using antithetic variables: ',paste(round(100*(var(estimator_raw)-var(estimator_anti))/var(estimator_raw),2),"%", sep = "")))
```

**Antithetic variable method** gives greater variance reduction.   

### Problem 2
Repeat 1, this time assuming the **move to the front rule** is being used.

#### 2. (a)
```{r}
X={}
Y={}
for (j in 1:n){
  result=list_problem_direct(p=p, elements=elements,list_initial=list_initial,
                             rule="front",var_reduce="control")
  X=c(X,result[1])
  Y=c(Y,result[2])
}
estimator_raw=X #raw estimator
cat(paste('Mean of raw estimator: ', mean(estimator_raw), 
          '\nVariance of raw estimator: ',var(estimator_raw)))
```

Using **control variables** method:
```{r}
c=-cov(X,Y)/var(Y)
estimator_control=X+c*(Y-50) #estimator using move to the front rule, control variable method
cat(paste('Mean of estimator using control variables: ', mean(estimator_control), '\nVariance of estimator using control variables: ',var(estimator_control)))
```
#### 2. (b)
```{r}
X={}
X_anti={}
for (j in 1:(n/2)){
  result=list_problem_direct(p=p, elements=elements,list_initial=list_initial,rule="front",var_reduce="antithetic")
  X=c(X,result[1])
  X_anti=c(X_anti,result[2])
}
estimator_anti=(X+X_anti)/2 #estimator using move to the front rule, antithetic variable method
cat(paste('Mean of estimator using antithetic variables: ', mean(estimator_anti), '\nVariance of estimator using antithetic variables: ',var(estimator_anti)))
```

#### 2. (c)   
```{r}
cat(paste('Variance reduction using control variables: ', paste(round(100*(var(estimator_raw)-var(estimator_control))/var(estimator_raw),2),"%", sep = "" ), '\nVariance reduction using antithetic variables: ',paste(round(100*(var(estimator_raw)-var(estimator_anti))/var(estimator_raw),2),"%", sep = "")))
```

**Antithetic variable method** gives greater variance reduction.


### Problem 3        
Here is another way we can try to use variance reduction. In each run start by reordering the p_i  so that  p_1 > p_2 > ... > p_{10}

in each period use U to determine the element requested as follows:

  if  U < p_1  request is for 1

  if p_1 < U < p1 + p_2  request is for 2

  and so on 

Assume **interchange rule** is used and find  (a), (b), (c) as in  Problem 1                        
                                                                      
                                                                      
#### 3. (a)                             

```{r}
X={}
Y={}
for (j in 1:n){
  result=list_problem_reorder(p=p, elements=elements,list_initial=list_initial,rule="interchange",var_reduce="control")
  X=c(X,result[1])
  Y=c(Y,result[2])
}
estimator_raw=X #raw estimator(reordering)
cat(paste('Mean of raw estimator: ', mean(estimator_raw), '\nVariance of raw estimator: ',var(estimator_raw)))

```

Using **control variables** method:
```{r}
c=-cov(X,Y)/var(Y)
estimator_control=X+c*(Y-50)
#estimator (reordering), using interchange rule, control variable method
cat(paste('Mean of estimator using control variables: ', mean(estimator_control), '\nVariance of estimator using control variables: ',var(estimator_control)))
```

#### 3. (b)                   

Using **antithetic variables** method:
```{r}
X={}
X_anti={}
for (j in 1:(n/2)){
  result=list_problem_reorder(p=p, elements=elements,list_initial=list_initial,rule="interchange",var_reduce="antithetic")
  X=c(X,result[1])
  X_anti=c(X_anti,result[2])
}
estimator_anti=(X+X_anti)/2
#estimator (reordering), using interchange rule, antithetic variable method
cat(paste('Mean of estimator using antithetic variables: ', mean(estimator_anti), '\nVariance of estimator using antithetic variables: ',var(estimator_anti)))
```
#### 3. (c)
```{r}
cat(paste('Variance reduction using control variables: ', paste(round(100*(var(estimator_raw)-var(estimator_control))/var(estimator_raw),2),"%", sep = "" ), '\nVariance reduction using antithetic variables: ',paste(round(100*(var(estimator_raw)-var(estimator_anti))/var(estimator_raw),2),"%", sep = "")))
```

**Antithetic variable method** gives greater variance reduction.

### Problem 4    
Repeat 3, this time assuming the **move to the front rule** is being used.

#### 4. (a)
```{r}
X={}
Y={}
for (j in 1:n){
  result=list_problem_reorder(p=p, elements=elements,list_initial=list_initial,rule="front",var_reduce="control")
  X=c(X,result[1])
  Y=c(Y,result[2])
}
estimator_raw=X #raw estimator(reordering)
cat(paste('Mean of raw estimator: ', mean(estimator_raw), '\nVariance of raw estimator: ',var(estimator_raw)))

```
    
Using **control variables** method:
```{r}
c=-cov(X,Y)/var(Y)
estimator_control=X+c*(Y-50)
#estimator (reordering), using move to the front rule, control variable method
cat(paste('Mean of estimator using control variables: ', mean(estimator_control), '\nVariance of estimator using control variables: ',var(estimator_control)))
```
#### 4. (b)
Using **antithetic variables** method:
```{r}
X={}
X_anti={}
for (j in 1:(n/2)){
  result=list_problem_reorder(p=p, elements=elements,list_initial=list_initial,rule="front",var_reduce="antithetic")
  X=c(X,result[1])
  X_anti=c(X_anti,result[2])
}
estimator_anti=(X+X_anti)/2
#estimator (reordering), using move to the front rule, antithetic variable method
cat(paste('Mean of estimator using antithetic variables: ', mean(estimator_anti), '\nVariance of estimator using antithetic variables: ',var(estimator_anti)))
```
#### 4. (c)
```{r}
cat(paste('Variance reduction using control variables: ', paste(round(100*(var(estimator_raw)-var(estimator_control))/var(estimator_raw),2),"%", sep = "" ), '\nVariance reduction using antithetic variables: ',paste(round(100*(var(estimator_raw)-var(estimator_anti))/var(estimator_raw),2),"%", sep = "")))
```

**Antithetic variable method** gives greater variance reduction.

### Problem 5
*What conclusions can you draw about which variance reduction technique works best in this problem?*

1. Reordering method doesn't help with variance reduction as much as method in the Ex 9i text does.      
2. Antithetic variable method better helps with variance reduction than control variables method in all above scenarios.        
3. Move to the front rule in problem 2 works better than interchange rule in problem 1. 
                                    
                                    
**In conclusion, the best technique to reduce variance in this problem is what we tried in problem 2(b), which is:  **      
***'Method in Ex 9i text without reordering' + 'Move to the front rule' + 'Antithetic variables method'.***  
